---
title: "Multivariate Viz"
---

```{r}
#Load data
education <- read.csv("https://mac-stat.github.io/data/sat.csv")

# Load libraries
library(ggplot2)
library(viridis)
library(viridisLite)
library(tibble)
library(gplots)
```

```{r, fig.width = 8, fig.height = 15}

# Remove the "State" column and use it to label the rows
# Then scale the variables
plot_data <- education |> 
  column_to_rownames("State") |> 
  data.matrix() |> 
  scale()

# Construct the heatmap
heatmap.2(plot_data,
  dendrogram = "none",
  Rowv = TRUE,            
  scale = "column",
  keysize = 0.7, 
  density.info = "none",
  col = hcl.colors(256), 
  margins = c(10, 20),
  colsep = c(1:7), rowsep = (1:50), sepwidth = c(0.05, 0.05),
  sepcolor = "white", trace = "none",
  main = "SAT data heatmap in 1994-95",
  alt = "A heatmap titled 'SAT data heatmap in 1995-95' which lists the states along the x-axis and 'frac, fracCat, expend, salary, ratio, verbal, sat, and math' as variables along the x axis. The colors represent the Z-score of each variable for each state. Particular areas with high Z scores are the expenditure variable for New York and New Jersey, and the ratio variable for Utah and Californa.",
  caption = "This data comes from the SAT dataset originally published in the Journal of Statistics Education by Deborah Lynn Guber. It shows state-by-state statistics for a number of SAT parameters and school spending."
)
```

Although this plot has many more than four variables, I chose it to be in my best work folder because it does a very good job showing areas of outliers and to which attention should be paid for many continuous numerical variables. This heat map does require some knowledge of the x axis titles, as well as an understanding that Z-score is the distance of a value from data points of the same parameter. I wish I knew how to change the x-axis titles to be more accessible, but the color palette in existence is accessible from a physical point of view, as it is high contrast. From looking at this graph we can clearly see the `ratio` variable (ratio of pupils to teacher) is an outlier in Utah and California, and New York and New Jersey have particularly high `expend` (education expenditure per student). 
