---
title: "Solo Project: Libraries in Scotland Local Authority Areas"
format: html
---

Steps to follow:

1. Natural earth package to find state boundaries or boundaries within a non-US country, or search internet for "__ shapefile" of a specific area of interest.

2. Find dataset that corresponds with country or state that will be used to color within boundaries.

3. Use sf package to draw boundaries in a plot.

4. Construct a chloropleth map of the variable of interest for the data set on the boundaries applied.


Shapefile for Scotland local authorities: 
https://www.spatialdata.gov.scot/geonetwork/srv/api/records/1cd57ea6-8d6e-412b-a9dd-d1c89a80ad62
(Under API pub-las section, download the shape-zip ZIP file)

CSV file for libraries in Scotland:
https://data.spatialhub.scot/dataset/libraries-is/resource/570de9e0-999e-4bfd-9ed6-71b09545965e
(Download button, and select CSV file, only using stationary libraries data and not "mobile libraries" data)


```{r load packages}
library(gplots)
library(rnaturalearth)
library(sf)
library(sfd)
library(readr)
```


```{r}
library_data <- read_csv("..Libraries_-_Scotland.csv")
```

```{r shapefile}


```






